Automatically Splitting a Two-Stage Lambda Calculus

### 2015-10-17 paper/examples.tex
various revisions

1`fun shade (refl : pipeline,
           albedo : pipeline) : pipeline =
  (fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl   (obj,next{2`xy`})`2`}
     * prev{`1`albedo (obj,next{2`xy`})`2`}`)
\end{lstlisting}
%
Whereas prior systems provide combinators like \texttt{par} as
language features\,\cite{Foley:2011} to support composition, \lang\ is
powerful enough to define those combinators in-language.

>>>>

1`fun shade (refl : pipeline, albedo : pipeline) : pipeline =
  fn (obj : object, next{2`xy`} : $2`coord`) =>
       2`prev{`1`refl (obj,next{2`xy`})`2`} * prev{`1`albedo (obj,next{2`xy`})`2`}`
\end{lstlisting}
%
Whereas prior systems provide combinators like \texttt{par} as
language features\,\cite{Foley:2011} to support composition, \lang\ is
powerful enough to define those combinators in-language.

### 2015-10-17 paper/relatedwork.tex
various revisions

However unlike splitting, where the goal is to emit code where first stage results
are computed once and then reused in multiple invocations of second
stage execution, the second (dynamic) function produced by binding time separation
does not use the results of the first; instead,
%only serves as a preprocessing step for the subsequent
%specialization phase of partial evaluation.
it has access to the first stage inputs and recomputes all required
first stage computations. As noted by Knoblock and
Ruf\,\cite{knoblock96}, it may be possible to modify the program
bifurcation algorithm to cache and reuse the intermediate results, but
this was never attempted.

@@

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ in
small ways to account for the specific evaluation model that we
propose; however, the key difference between our work and this prior
work is that we instead focus on the problem of splitting.

@@

As noted in \ref{sec:graphics}, splitting algorithms have also been a
topic of interest in programming systems for computer graphics, where,
to achieve high performance, programs are manually separated into
components by frequency of execution corresponding to graphics
hardware pipeline stages. The software engineering challenges of
modifying multiple per-stage programs have led to suggestions of
writing graphics programs in an explicitly-staged programming
language\,\cite{Proudfoot:2001,Foley:2011,He:2014} and deferring the
task of pass separation to the compiler. However, all prior splitting
efforts in computer graphics, like that of Knoblock and
Ruf\,\cite{knoblock96} have been limited to simple, imperative
languages.

>>>>

However unlike splitting, where the goal is to emit code where first stage results
are computed once and then reused in multiple invocations of second
stage execution, the second (dynamic) function produced by binding time separation
does not use the results of the first; instead,
%only serves as a preprocessing step for the subsequent
%specialization phase of partial evaluation.
it has access to the first stage inputs and recomputes all required
first stage computations. As noted by Knoblock and
Ruf\,\cite{knoblock96}, it may be possible to modify the program
bifurcation algorithm to cache and reuse the intermediate results, but
this was never attempted.
Alternatively, our algorithm could potentially be used as the basis of a general
bifurcation algorithm in a partial evaluator.

@@

The type-system of \lang\ is closely related to this later line of
work on metaprogramming and staged computation.  The specific
extension to the typed lambda calculus that we use here is based on
the $\fut$ modality of Davies~\cite{DP01-modal}.  Our types differ 
in the restriction to two stages and the addition of $\curr$; 
however, the key difference between our work and this prior
work is that we instead focus on the problem of splitting.

@@

As noted in \ref{sec:graphics}, splitting algorithms have also been a
topic of interest in programming systems for computer graphics, where,
to achieve high performance, programs are manually separated into
components by frequency of execution corresponding to graphics
hardware pipeline stages. The software engineering challenges of
modifying multiple per-stage programs have led to suggestions of
writing graphics programs in an explicitly-staged programming
language\,\cite{Proudfoot:2001,Foley:2011,He:2014} and deferring the
task of pass separation to the compiler. However, all prior splitting
efforts in computer graphics, like that of Knoblock and
Ruf\,\cite{knoblock96}, have been limited to simple, imperative
languages.

\paragraph{Defunctionalization.}
Defunctionalization \cite{Reynolds72} is a program transformation
that eliminates high-order functions and replaces them with lower
order functions that manipulate a data structure encoding the original control flow.
This operation has similarity to our splitting transformation, 
which eliminates staging in a program by encoding control flow in a data structure
passed between the stage one and stage two outputs.
It would interesting to explore the possibility of a semantics-preserving transformation
to convert a staged program into a higher-order form,
and then applying defunctionalization to obtain our results.

### 2015-10-17 paper/semantics.tex
various revisions

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within $\next$ blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
and will be discussed in \ref{sec:needGround}.

@@

\paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code (compatibility rules) 
where it either performs a beta reduction (reduction rules) or marks the term as being evaluated (value marking rules). 
A ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a beta reduction or marks a value.
It also descends into $\pure$ and $\next$ blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, $\pure$ blocks of ground values, and $\next$ blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
expensive computation $e$ depending on how many times $x$ appears in $e$. 
Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}\end{abstrsyn}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
We read this as saying somewhere within $e$ there was the subterm $q$ which needs to be hoisted out,
yielding the new term $e'$ which has a new variable $y$ where the $q$ used to be.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a top-level program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

@@

The statics and dynamics of \lang\ are related by the type safety theorems below, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
Note how the progress theorem for world \bbonem\ states that every well-typed term must take 
either a standard step or a hoisting step.

\begin{abstrsyn}
\begin{theorem} [Progress] \leavevmode
\label{thm:progress} 
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem} [Preservation] \leavevmode
\label{thm:preservation} 
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, then $\typestwo q B$ and $\typesone [\Gamma,\coltwo y B] {e'} A$.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}
\end{abstrsyn}

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let fun qs k = 
      prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let fun qs k = 
	  let val y = ... in y
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a large nest of case statements open on $k$.

>>>>

We express two-stage programs as terms in \lang, a typed, modal lambda calculus. 
Although \lang\ describes computations that occur in two stages,
we find it helpful for the specification of splitting to differentiate terms into three \emph{worlds}.
A world is essentially a slightly finer classification than a stage.
Whereas there is only one world, \bbtwo, for second-stage computations,
there are two worlds corresponding to the first stage:
\bbonem\ for \emph{mixed} first-stage computations, which may contain second-stage subterms within {\tt next} blocks,
and \bbonep\ for \emph{ground} first-stage computations, which may not.
The distinction between these two first-stage worlds is necessary for an efficient splitting algorithm
and will be discussed in \ref{sec:needGround}.

@@

\paragraph{World \bbonep.} Since the ground fragment of the language is not dependent on other worlds, the semantics of ground is just that of a monostage language.
Thus, $\steppure e {e'}$ traverses into subterms to to find the left-most unevaluated code where it performs a reduction. 
A ground value $u$ comprises only units, injections, tuples, and functions, where the body of the function may be any ground term.

\paragraph{World \bbonem.}
Like its ground counterpart, the \bbonem\ step judgment, $\stepmix e {e'}$, finds the left-most unevaluated subterm and performs a reduction.
It also descends into {\tt gr} and {\tt next} blocks, transitioning into one of the other two steps ($\stepsym \bbonep$ or $\stepsym \bbtwo$) there. 
The value form for \bbonem, called a partial value, comprises units, tuples, functions, {\tt gr} blocks of ground values, and {\tt next} blocks containing \emph{only a stage two variable}.
This strong requirement ensures that second-stage computations are not duplicated 
when partial values are substituted for a variable.
This is a departure from the staged semantics of \cite{taha-thesis-99} and \cite{davies96}.
Whereas those semantics interpret values of type $\fut A$ to mean ``code of type $A$ that can be executed in the future,''
ours interprets $\fut A$ to mean ``a reference to a value that will be accessible in the future.''
This contrast stems from differing goals: metaprogramming explicitly intends to model code manipulating code,
whereas our applications feel more natural with an eager interpretation of {\tt next}.
One consequence of the stronger requirement on partial values is that a new kind of step is necessary to put terms into that form.
To illustrate, consider:
\begin{lstlisting}
2`prev{`1`(fn x : $int => e') (next{`2`e`1`})`2`}`
\end{lstlisting}
We could reduce this to $\prev{[\next e/x]e'}$, but this may potentially duplicate an
expensive computation $e$ depending on how many times $x$ appears in $e'$. 
Instead, we choose to \emph{hoist} $e$ outside, binding it
to a temporary variable $y$, and substituting that variable instead:
\begin{lstlisting} 
2`let val y = e in prev{`1`[next{`2`y`1`}/x]e'`2`}`
\end{lstlisting}\end{abstrsyn}
This behavior is implemented by the $\lift e {e'}$ judgment, called a \emph{hoisting step}.
We read this as saying that somewhere within $e$ there was a subterm $q$ which needs to be hoisted out,
yielding the new term $e'$ which has a new variable $y$ where $q$ used to be.
These steps occur when a \next\ block has contents that are a residual but (to prevent loops) not when those contents are already a variable.
In essence, the rules for hoisting steps operate by ``bubbling up'' a substitution to the innermost containing \prev,
where it is reified into a {\tt let} statement.
\footnote{Because a program is a term at \bbtwo, this {\tt prev} always exists.
Otherwise, the semantics would need a mechanism to accumulate the bindings that hoisting steps create.}

Consider the following example, where {\tt P} has type $\curr(\rmunit + \rmunit)$,\begin{lstlisting}
           1`caseg P of _ => next{`2`0`1`} | _ => next{`2`literalone`1`}`
\end{lstlisting}
Depending on what {\tt P} evaluates to in the first stage, whole term will step to either {\tt next\{0\}} or {\tt next\{1\}}.
In this sense, we can see {\tt case} (at world \bbonem) and {\tt caseg} as the constructs that facilitate
all cross-stage communication.

@@

The statics and dynamics of \lang\ are related by the type safety theorems below, again annotated by world.
In all cases, $\Gamma$ may be any list of variable bindings at world \bbtwo,
representing the second stage binders under which we are evaluating.
Note how the progress theorem for world \bbonem\ states that every well-typed term must take 
either a standard step or a hoisting step.
\begin{abstrsyn}
\begin{theorem} [Progress] \leavevmode
\label{thm:progress} 
\begin{itemize} 
\item If $\typesone e A$, then either $e$ has the form $\exv v$, or $\stepmix e {e'}$, or $\lift e {e'}$.
\item If $\typespure e A$, then either $e$ has the form $\exv u$, or $\steppure e {e'}$.
\item If $\typestwo e A$, then either $e$ has the form $\exv q$, or $\steptwo e {e'}$.
\end{itemize}
\end{theorem}
\begin{theorem} [Preservation] \leavevmode
\label{thm:preservation} 
\begin{itemize} 
\item If $\typesone e A$ and $\lift e {e'}$, \\
	then $\typestwo q B$ and \mbox{$\typesone [\Gamma,\coltwo y B] {e'} A$}.
\item If $\typeswor e A$ and $\stepwor e {e'}$, then $\typeswor {e'} A$.
\end{itemize}
\end{theorem}
\end{abstrsyn}

@@

We can also directly observe the reuse of first-stage computations across
multiple second-stage computations. To run many order statistics queries
$\texttt{k1},\dots,\texttt{km}$ on the same list, run the program:
\begin{lstlisting}
2`let fun qs k = prev{`1`qss (`3`gr{`1`[7,4,2,5,9,...,3]`3`}`1`,next{`2`k`1`})2`}`
2`in (qs k1,...,qs km)`
\end{lstlisting}
To see that the partial application of \texttt{qss} is only evaluated once,
observe that after the first stage of evaluation, the residual has the form:
\begin{lstlisting}
2`let fun qs k = (let val y = ... in y)
2`in (qs k1,...,qs km)`
\end{lstlisting}
where $y$ is bound to a large nest of case statements open on $k$.

### 2015-10-17 paper/splitting.tex
various revisions

Since direct staged term evaluation (\ref{sec:semantics})
reduces all first stage terms to value forms without any remaining stage two work,
the distinction between \bbonem\ and \bbonep\ is unnecessary.
In contrast, when performing program transformations before the first stage inputs are known,
it is valuable to form a clear distinction between ground and mixed terms.
This was similarly observed in prior work seeking to implement self-applicable partial evaluators\,\cite{Mogensen89a,Mogensen89b}.
While this paper assumes that ground annotations already exist in the input,
it may be possible to use binding-time analysis techniques to automatically insert them.
Ground is also similar to the validity mechanism in ML5\,\cite{murphy05}.
\end{abstrsyn}

>>>>

Since direct staged term evaluation (\ref{sec:semantics})
reduces all first stage terms to value forms without any remaining stage two work,
the distinction between \bbonem\ and \bbonep\ is unnecessary.
In contrast, when performing program transformations before the first stage inputs are known,
it is valuable to form a clear distinction between ground and mixed terms.
This was similarly observed in prior work seeking to implement self-applicable partial evaluators\,\cite{Mogensen89a,Mogensen89b}.
While this paper assumes that ground annotations already exist in the input,
it may be possible to use binding-time analysis techniques to automatically insert them.
Ground is also similar to the validity mechanism in ML5\,\cite{murphy05}.

\subsection{Typing the Boundary Data Structure}

One of the central features of our splitting algorithm is that
it encodes the control flow behavior of the original 
staged program into the boundary data structure.
For instance, the {\tt case} and {\tt caseg} splitting rules 
put injection tags on the boundary based on which branch was taken,
and the {\tt fn} rule adds a roll tag to the boundary.
As a result, the boundary value passed between the two output programs has a (potentially recursive) structure like a tree or list.

This structure can be described with a type; for instance, the staged quickselect yields a binary search tree.
Indeed, for most of our examples in \ref{sec:examples}, inferring this type is straight-forward. 
However, we do not yet have a formal characterization of these boundary types that is total for all \lang\ programs.
\footnote{Observe that in the outputs of splitting we could have ommitted the {\tt roll} tag from functions or 
replaced $\inl a$ and $\inr b$ with $\tup {0,a}$ and $\tup {1,b}$, and the proof of correctness would still have gone through.
We chose the tags, however, in order to keep the typed interpretation more natural, even in absence of a formal result.
}
\end{abstrsyn}
